Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. At any point, we may choose to \textit{flush} some
of the writes in buffer to the global pomset of events.
\TODO{\(\mc{A}_{\ms{TSO}}\), \(\mc{A}_b\)}
of type \(\ms{Buf} = \mc{A}_b^*\)
of lists of write actions, which we will interpret as linear pomsets ordered by
index. In particular, we define the monad \(\ms{TSO} =
\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\). 
\TODO{using \(\mc{A}_{\ms{TSO}}\)}
Given a buffer \(\ms{Buf}\), we can define the
result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! 
\TODO{segue}
This is implemented via the morphism (called
\(\ms{split}\) in \cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
\TODO{reflow}
On the other hand, reads events will first attempt to read from the buffer, and,
if there is no corresponding write in the buffer, will read an arbitrary value,
in both cases pushing an event to the global pomset; that is, the semantics of
reads and writes are given by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
\TODO{text explaining this}
\TODO{in the presence of parallel composition}

>>>>

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. At any point, we may choose to \textit{flush} some
of the writes in buffer to the global pomset of events. In particular, we will
introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write
actions}, where an action \(\bufloc{x} := v\) by a thread denotes adding a write
\(x := v\) to the thread's write buffer. We may then define the set of TSO
actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be
defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will
interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In
particular, we define the monad \(\ms{TSO} =
\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where \(\Sigma\) is the stream
action of finite pomsets over \(\mc{A}_{\ms{TSO}}\) on pomsets over
\(\mc{A}_{\ms{TSO}}\); we can view \(\ms{PO}\) as a submonad of this monad in
the standard manner.

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! 
\TODO{segue}
This is implemented via the morphism (called
\(\ms{split}\) in \cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
\TODO{reflow}
On the other hand, reads events will first attempt to read from the buffer, and,
if there is no corresponding write in the buffer, will read an arbitrary value,
in both cases pushing an event to the global pomset; that is, the semantics of
reads and writes are given by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
\TODO{text explaining this}
\TODO{in the presence of parallel composition}
